\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2016-12-28 document/XSGRB_sample.tex
Writing completeness section and adding figures

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sample selection criteria} \label{samplecrit}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Being of transient nature, it is difficult to impose strong sample selection
criteria on GRBs, without hampering the follow-up effort. Many natural follow-up
restrictions exists from already, being it weather conditions, pointing
restrictions of the telescope or poorly localized bursts as reported by the
\textit{Swift}-telescope. To maximize the return of the follow-up campaign
we have chosen a few selection criteria that attempts to provide an unbiased
selection of bursts, while allowing for a high success-rate

\footnote{Note that in the P84 proposal the criteria have
been stated a bit differently, the visibility constraint being replaced by a
declination + Sun angle constraint. The above criteria are however those
defining the sample.}

@@

A significant fraction of the bursts presented here \todo{Insert exact
number} have already had their hosts investigated in \citet{Kruhler2015}, for
which extractions of the hosts exist. The focus of the data presented here are
on the afterglows themselves and in the absence of a clear afterglow, the host.
We will not, however, investigate the hosts.

@@

All the observations in this sample has been carried out using the
cross-dispersed echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on
two of the four Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal)
during the duration of this follow-up campaign, which covers the entire lifetime
of X-shooter. Observations have been carried out from the ESO period P84
through P98 under the following programme IDs: 098.A-0055, 097.A-0036, 096.A-0079,
095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260.

The first GRB followed up is GRB090313, observed 15th of March 2009, which was
while X-shooter was mounted on UT3 during the commissioning of the instrument.
The bursts observed during the commisioning period is not a part of our
statistical sample, but are nonetheless published as part of the bursts observed
by the X-shooter GRB team. The first burst observed after science verification
was completed when X-shooter was moved to UT2 is GRB091018 which thereby
constitute the first burst entering our statistical sample. For all bursts that
fulfill our sample selection criteria, described in Sect. \ref{samplecrit},
spectroscopic follow-up has been attempted with X-shooter. Various conditions
can affect our ability to follow up a given burst, and a discussion of this
effect is included in Sect. \ref{badbursts}.

@@

In this section we describe the efficiency of the follow-up effort and the characteristics of the observed bursts. We also assess the degree to which the sample obtained is complete.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Follow-up timing and afterglow brightness}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A rapid follow-up is essential for the accurate designation of GRB the origin. The redshift determination of burst where the host is not easily

\begin{figure}
	\centerline{\includegraphics[width=8cm]{figures/timing.pdf}}
	\caption{Follow-up timing and afterglow magnitude at the start of observation. The points have been colored based on the redshift of the corresponding burst. }
	\label{fig:timing}
\end{figure}

@@

Based on the sample criteria specified in Sect. \ref{samplecrit}, a total of 152 \todo{Get number} bursts has been triggered by \textit{Swift} in the period since the comissioning of VLT/X-shooter in \todo{Get date}. This constitutes the \textit{statistical sample}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{On the nature of triggered GRBs without detected optical afterglow.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

@@

Due of the stochastic nature of the \lya-forest, the blue wing of the Lyman
alpha absorption line is randomly superposed with forest systems, along with
strong absorption from \mnii~and \SIiii, making it notoriously difficult to
model. Additionally, Similarly so, the red wing has got the ISM signatures
imprinted on it, especially strong absorption due to \SIii, \sii and \nv~which
can exhibit significant velocity structure. Along with instrumental effects, the
generative model for the data that we would use in a likelihood-based analysis
would be almost boundless in complexity, making formal fitting of the column
densities outside the scope of this paper.

@@

\subsection{GRB100418A (z=0.624)}
The data presented here also formed the basis of GCN \#
10620\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10620.gcn3}} and GCN \#
10631\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10631.gcn3}} and is published
in \citet{DeUgartePostigo2011}. The burst have been followed up in three epochs
of observations, 0.4, 1.4, and 2.4 days after the burst, each lasting 4.8 ks.
The unambiguous redshift of the host, $z=0.624$, is found from the simultaneous
detection of emission features belonging to nebular lines, including \hi, \oii,
\oiii, \neiii, \nii, \sii, \siii, and \hei~as well as absorption features due
to the presence of \znii, \crii, \feii, \mnii, \mgii, \mgi, \tiii, and \caii,
all at a consistent redshift. Temporal evolution of the fine-structure lines
belonging to \feii$^*$ is found between the epochs.

@@

\subsection{GRB120815A* (z = 2.358)} 
Not a part of the statistical sample, thisburst also formed the basis of  GCN
\# 13649\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/13649.gcn3}} and is
published in \citet{Kruhler2013}. Observations started 1.69 hours after the BAT
trigger and consist of 2.4 ks integration. A bright afterglow continuum is
detected across the entire spectral coverage of X-shooter, with a multitude of
absorption lines superposed. Absorption features from the host at $z = 2.358$
include a DLA as well as metal absorption lines from \nv, \sii, \SIii, \oi,
\civ, \SIiv, \feii, \alii, \aliii, \mnii, \mgii, and \mgi. Additionally
fine-structure lines from \NIii and \feii are exited local to the GRB.
Intervening systems are found at $z = 1.539$, $z = 1.693$, and $z = 2.00$.

\subsection{GRB120909A (z = 3.929)}
The data presented here has formed the basis of GCN \#
13730\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/13730.gcn3}}, but is not
published otherwise. A very rapid follow-up, starting only 1.7 hours after the
BAT trigger, this 1.2 ks observation captures a very bright afterglow
continuum, starting at 4500 \AA, signifying the onset of the Lyman limit for a 
system at $z = 3.929$. Absorption from high-column density hydrogen leaves very
prominent absorption features in the form of \lya, \lyb, and \lyg, visible in
the Lyman alpha forest. Metal absorption lines arising from \feii, \NIii,
\SIii, \sii, \alii, \aliii, \cii, \oi, \civ, and \znii~are all detected along
with the corresponding fine-structure lines from (\feii$^*$, \SIii$^*$,
\oi$^*$, \oi$^**$, \cii$^*$), securely anchoring the redshift of the host.

@@

\subsection{GRB121024A (z = 2.300)}
The data presented here also formed the basis of GCN \#
13890\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/13890.gcn3}} and is published
in \citet{Friis2015}. Also rapid, starting 1.8 hours after the \textit{Swift}
trigger, a bright afterglow continuum is visible across all arms. A broad
absorption feature from Lyman alpha, along with narrow lines from \civ, \SIii,
\SIiv, \feii, \sii, and \alii, as well as fine-structure lines associated with
\SIii$^*$ are all detected at $z = 2.300$, securely setting it as the redshift
of the GRB.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sample selection criteria} \label{samplecrit}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Being of transient nature, it is difficult to impose strong sample selection
criteria on GRBs, without hampering the follow-up effort. Many natural follow-up
restrictions exists from already, being it weather conditions, pointing
restrictions of the telescope or poorly localized bursts as reported by the
\textit{Swift}-telescope. To maximize the return of the follow-up campaign
we have chosen a few selection criteria that attempts to provide an unbiased
selection of bursts, while allowing for a high success-rate

\footnote{Note that in the P84 proposal the criteria have
been stated a bit differently, the visibility constraint being replaced by a
declination + Sun angle constraint. The below criteria are however those
defining the sample.}

@@

A fraction of the bursts presented here \todo{Insert exact
number} have already had their hosts investigated in \citet{Kruhler2015}, for
which extractions of the hosts exist. The focus of the data presented here are
on the afterglows themselves and in the absence of a clear afterglow, the host.
We will not, however, investigate the hosts.

@@

Observations in this sample has been carried out using the cross-dispersed
echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on two of the four
Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal) during the duration
of this follow-up campaign, which covers the entire lifetime of X-shooter.
Observations have been carried out from the ESO period P84 through P98 under the
following programme IDs: 098.A-0055, 097.A-0036, 096.A-0079, 095.A-0045,
094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088, 089.A-0067,
088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260. A few additional
bursts, not a part of the \textit{statistical sample}, have been included from
092.D-0056(A) (PI: Rau), 092.D-0633(E) (PI: Greiner), and Levan under
095.B-0811(B) (PI: Levan). This sample represents \textit{all} GRBs that have
ever been followed up by X-shooter.

The first GRB followed up is GRB090313, observed 15th of March 2009, which was
while X-shooter was mounted on UT3 during the commissioning of the instrument.
The bursts observed during the commissioning period is not a part of our
statistical sample, but are nonetheless published as part of the bursts observed
by the X-shooter GRB team. The first burst observed after science verification
was completed when X-shooter was moved to UT2 is GRB091018 which thereby
constitute the first burst entering our statistical sample. For all bursts that
fulfill our sample selection criteria, described in Sect. \ref{samplecrit},
spectroscopic follow-up has been attempted with X-shooter. Various conditions
can affect our ability to follow up a given burst, and a discussion of this
effect is included in Sect. \ref{badbursts}.

@@

In this section we describe the efficiency of the follow-up effort and the
characteristics of the observed bursts. We also assess the degree to which the
sample obtained is complete. An important note is that here we present
\textit{all} GRBs that have ever been observed with X-shooter, while only a
subset of these constitute our \textit{statistical} sample. The statistical
sample is based on the selection criteria described in Sect. \ref{samplecrit}
and should represent an unbiased sample of the underlying, parent GRB
population. Single bursts outside the sample criteria have been followed up on
the basis of the curiosity of their light curve or their brightness, but is not
discussed as part of the investigation of the GRB population. A prime example of
a burst outside the statistical sample is GRB161023A with it's rich
spectrum containing 11 intervening absorption systems and pending further analysis.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Follow-up timing and afterglow brightness} \label{timing}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Redshift determination of bursts where the host is too faint for a spectroscopic
redshift relies on the absorption spectrum imprinted on the GRB afterglow
continuum, which again depends on the afterglow brightness at time of follow-up.
Because the afterglow rapidly fades \citep{Nousek2006, Vecchio2016}\todo{Better references} a rapid
follow-up is essential for the accurate designation of the GRB host \todo{Get
reference}. In Fig. \ref{fig:timing} we plot the follow-up delay from the BAT
trigger to the start of the spectroscopic observation. Visible as the
shortest delays are the bursts observed in RRM-mode in which where the fastest follow-up
between BAT trigger and start of spectroscopic observations is for the short, $z = 1.717$, GRB160410A which was
followed up only 8.4 minutes after the BAT trigger.

\begin{figure}
	\centerline{\includegraphics[width=8cm]{figures/timing.pdf}}
	\caption{Follow-up timing and afterglow magnitude at the start of observation.
	The points have been colored based on the redshift of the corresponding burst.
	Points with black faces does not have a redshift determination and downarrows
	indicate a non-detection in the acquisition image. In blue is shown the
	fractional redshift completess as a function of delay-time.}
	\label{fig:timing}
\end{figure}

To illustrate the effect of redshift completeness loss due to late follow-up, we
overplot the redshift completeness fraction over the delay time. Visible is the
increasing fraction of GRBs without a good redshift determination for increasing
delays. That the redshift completeness of the bursts we have followed up are $>$
90 percent \todo{Get number}, shows the ability of VLT/X-shooter to successfully
get redshifts. Not shown in the figure is an additional 12 \todo{Get number}
bursts that we have redshift determinations from late-time host observations
with delay-times longer than $\sim$10 days.

@@

Based on the sample criteria specified in Sect. \ref{samplecrit}, a total of 158
\todo{Get number} bursts has been triggered by \textit{Swift} in the period
since the commissioning of VLT/X-shooter in \todo{Insert date}. This sample
constitutes the \textit{statistical sample} on which we will try to address the
GRB population properties. Of this sample, 86 \todo{Get number} have been
spectroscopically followed up with X-shooter. In order to assess whether our
sample criteria and the subset of bursts followed up represent the underlying
parent population of GRB, we compare some intrinsic GRB properties of our sample
to the full sample of GRBs followed up by \textit{Swift}. Under the assumption
that \textit{Swift} randomly samples the underlying GRB population, if the
sample statistics of our subsample are similar to the \textit{Swift}-sample, we
can assume that the population properties should be conserved. There is some
evidence that the burst duration does not scale with redshift as expected
through time-dilation\citep{Kocevski2013, Littlejohns2013a}, pointing towards
that we could be missing an increasing fraction of the bursts at higher
redshifts, but \citet{Lien2016} argues that we should only be missing the bright
"short" GRBs", but to the degree that \textit{Swift} is complete, we can test if
the sample presented here is complete. We show the comparison between the burst
properties in Fig. \ref{fig:swift_complete}.

\begin{figure*}
	\centerline{\includegraphics[width=18cm]{figures/completeness_BAT.pdf}}
	\caption{Comparison between bursts properties of all bursts observed with
	\textit{Swift}, the properties of the ones that fullfill the sample criteria
	specified in Sect. \ref{samplecrit}, and the subset that has been observed as
	part of the statistical sample. The left plot shows the fluence in 15-150 keV
	as observed by BAT. The middle panel shows the 0.3 - 10 keV flux, 11 hours
	after the bursts as measured by XRT. Based on the XRT spectrum,
	\citep{Evans2009}, a hydrogen column density has been inferred which we show in
	the right most panel.}
	\label{fig:swift_complete}
\end{figure*}

Using the observational characteristics of the 1244 \todo{Get number} bursts
observed by \textit{Swift} to date \cite{Gehrels2009}, and the X-ray derived
hydrogen column densities \citep{Evans2009}, we can quantify the degree to which
our sample is biased against the overall \textit{Swift} sample. The values are
queried from then Three main samples are of interest to compare to assess the
completenessof the follow-up campaign; the full \textit{Swift} sample consisting
of all the burst observed by \textit{Swift} (1), the burst that fulfill the
selection criteria imposed in Sect. \ref{samplecrit} (2), and the bursts
actually followed up with X-shooter (3).

\input{tables/sample_properties.tex}

Using a
2-sided Kolmogorov-Smirnov (KS), we can assess the degree to which the null that
the two distribution are drawn from the same distribution can be rejected.

\begin{figure}
	\centerline{\includegraphics[width=9cm]{figures/p_values.pdf}}
	\caption{Relational graph shown the .}
	\label{fig:p_values}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The nature of triggered GRBs without detected optical afterglow.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{On the redshift distribution of GRBs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Because the redshift completeness of our statistical sample is  54 percent
\todo{Get number}, making inference of the underlying redshift distribution of
GRBs is difficult. Because a large fraction of the bursts for which we were
unable to secure a redshift are missing due to terrestrial reasons, see Sect.
\ref{badbursts}, imposing additional unbiasing cuts on the sample can increase
the redshift completeness at the cost of sample size. As long as the cuts we
impose are not selective in terms of the burst properties, the sample properties
stay the same. The first cut we try is telescope availability such as bad
weather, visitor rejected the TOO, or unavailability of the instrument. Because
telescope availability is independent on the burst properties, by cutting away
these from our sample, the underlying redshift distribution should be conserved.
By trimming away the 30 bursts \todo{Get number} that can be rejected due to
telescope availability, the redshift completeness increases to 63 percent
\todo{Get number}. As is shown in Sect. \ref{timing}, the redshift completeness
decreases with increasing follow-up delay, we can additionally increase the
redshift completeness by imposing cuts on the follow-up delay. Our ability to
rapidly follow up a burst is also not dependent on the burst properties and
therefore we can additionally cut delays longer than 24 hours (all burst in the
sample are in principle observable within 24 hours).

@@

The biased introduced in the redshift distribution by the increased ability to
secure redshift of optically brighter bursts has been investigated by
\citet{Turpin2016} which find that we systematically find redshift for longer
GRBs. Additionally only the brightest GRBs are seen above redshift $z \gtrsim
1$. \todo{Additional text}

@@

Due of the stochastic nature of the \lya-forest, the blue wing of the Lyman
alpha absorption line is randomly superposed with forest systems, along with
strong absorption from \mnii~and \SIiii, making it notoriously difficult to
model. Additionally, Similarly so, the red wing has got the ISM signatures
imprinted on it, especially strong absorption due to \SIii, \sii~and \nv~which
can exhibit significant velocity structure. Along with instrumental effects, the
generative model for the data that we would use in a likelihood-based analysis
would be almost boundless in complexity, making formal fitting of the column
densities outside the scope of this paper.

@@

\subsection{GRB100418A (z=0.624)}
The data presented here also formed the basis of GCN \#
10620\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10620.gcn3}} and GCN \#
10631\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10631.gcn3}} and is published
in \citet{DeUgartePostigo2011}. The burst have been followed up in three epochs
of observations, 0.4, 1.4, and 2.4 days after the burst, each lasting 4.8 ks.
The unambiguous redshift of the host, $z=0.624$, is found from the simultaneous
detection of emission features belonging to nebular lines, including \hi, \oii,
\oiii, \neiii, \nii, \sii, \siii, and \hei~as well as absorption features due
to the presence of \znii, \crii, \feii, \mnii, \mgii, \mgi, \tiii, and \caii,
all at a consistent redshift. Temporal evolution of the fine structure lines
belonging to \feii$^*$ is found between the epochs.

@@

\subsection{GRB120815A* (z = 2.358)} 
Not a part of the statistical sample, thisburst also formed the basis of  GCN
\# 13649\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/13649.gcn3}} and is
published in \citet{Kruhler2013}. Observations started 1.69 hours after the BAT
trigger and consist of 2.4 ks integration. A bright afterglow continuum is
detected across the entire spectral coverage of X-shooter, with a multitude of
absorption lines superposed. Absorption features from the host at $z = 2.358$
include a DLA as well as metal absorption lines from \nv, \sii, \SIii, \oi,
\civ, \SIiv, \feii, \alii, \aliii, \mnii, \mgii, and \mgi. Additionally
fine structure lines from \NIii and \feii are exited local to the GRB.
Intervening systems are found at $z = 1.539$, $z = 1.693$, and $z = 2.00$.

\subsection{GRB120909A (z = 3.929)}
The data presented here has formed the basis of GCN \#
13730\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/13730.gcn3}}, but is not
published otherwise. A very rapid follow-up, starting only 1.7 hours after the
BAT trigger, this 1.2 ks observation captures a very bright afterglow
continuum, starting at 4500 \AA, signifying the onset of the Lyman limit for a 
system at $z = 3.929$. Absorption from high-column density hydrogen leaves very
prominent absorption features in the form of \lya, \lyb, and \lyg, visible in
the Lyman alpha forest. Metal absorption lines arising from \feii, \NIii,
\SIii, \sii, \alii, \aliii, \cii, \oi, \civ, and \znii~are all detected along
with the corresponding fine structure lines from (\feii$^*$, \SIii$^*$,
\oi$^*$, \oi$^**$, \cii$^*$), securely anchoring the redshift of the host.

@@

\subsection{GRB121024A (z = 2.300)}
The data presented here also formed the basis of GCN \#
13890\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/13890.gcn3}} and is published
in \citet{Friis2015}. Also rapid, starting 1.8 hours after the \textit{Swift}
trigger, a bright afterglow continuum is visible across all arms. A broad
absorption feature from Lyman alpha, along with narrow lines from \civ, \SIii,
\SIiv, \feii, \sii, and \alii, as well as fine structure lines associated with
\SIii$^*$ are all detected at $z = 2.300$, securely setting it as the redshift
of the GRB.
